### 2017-07-07 abstract.tex
edits to inference

  Serializability is a well-understood correctness criterion that
  simplifies reasoning about the behaviour of concurrent transactions
  by ensuring they are \emph{isolated} from each other while they
  execute.  However, enforcing serializable isolation comes at a steep
  cost in performance because it necessarily restricts opportunities
  to exploit concurrency even when such opportunities would not
  violate application-specific invariants. As a result, database
  systems in practice support, and often encourage, developers to
  implement transactions using weaker alternatives. These alternatives
  break the strong isolation guarantees offered by serializable
  transactions to permit greater concurrency. Unfortunately, the
  semantics of weak isolation is poorly understood, and usually
  explained only informally in terms of low-level implementation
  artifacts. Consequently, verifying high-level correctness properties
  in such environments remains a challenging problem.

  To address this issue, we present a novel program logic that enables
  compositional reasoning about the behavior of concurrently executing
  weakly-isolated transactions.  Recognizing that the proof burden
  necessary to use this logic may dissuade application developers, we
  also describe an inference procedure based on this foundation that
  ascertains the weakest isolation level that still guarantees the
  safety of high-level consistency assertions associated with such
  transactions.  The key to effective inference is the observation
  that weakly-isolated transactions can be viewed as functional
  (monadic) computations over database state, allowing us to treat
  their operations as state transformers over the database.  This
  interpretation enables automated verification using off-the-shelf
  SMT solvers.

>>>>

  Serializability is a well-understood correctness criterion that
  simplifies reasoning about the behavior of concurrent transactions
  by ensuring they are \emph{isolated} from each other while they
  execute.  However, enforcing serializable isolation comes at a steep
  cost in performance because it necessarily restricts opportunities
  to exploit concurrency even when such opportunities would not
  violate application-specific invariants. As a result, database
  systems in practice support, and often encourage, developers to
  implement transactions using weaker alternatives. These alternatives
  break the strong isolation guarantees offered by serializable
  transactions to permit greater concurrency. Unfortunately, the
  semantics of weak isolation is poorly understood, and usually
  explained only informally in terms of low-level implementation
  artifacts. Consequently, verifying high-level correctness properties
  in such environments remains a challenging problem.

  To address this issue, we present a novel program logic that enables
  compositional reasoning about the behavior of concurrently executing
  weakly-isolated transactions.  Recognizing that the proof burden
  necessary to use this logic may dissuade application developers, we
  also describe an inference procedure based on this foundation that
  ascertains the weakest isolation level that still guarantees the
  safety of high-level consistency assertions associated with such
  transactions.  The key to effective inference is the observation
  that weakly-isolated transactions can be viewed as functional
  (monadic) computations over an abstract database state, allowing us
  to treat their operations as state transformers over the database.
  This interpretation enables automated verification using
  off-the-shelf SMT solvers.

### 2017-07-07 inference.tex
Edit.
edits to inference

Rely-Guarantee framework presented in the previous section facilitates
modular proofs for weakly-isolated transactions, but imposes a
non-trivial annotation burden on the programmer. In particular, it
requires each statement ($c$) of the transaction to be annotated with
a stable pre- ($P$) and post-condition ($Q$), and loops to be
annotated with stable inductive invariants ($\psi$). While weakest
pre-condition style predicate transformers can help in inferring $P$'s
and $Q$'s for regular statements, inferring inductive loop invariants
is still a challenge, even for the simpler form of loops we admit.
Thus, predicate transformer-based approach to proof automation is
unlikely to be effective in this case. We therefore present an
alternative approach to RG proof automation in this section that is
based on state transformers instead of predicate transformers. The
idea is to infer the effect that each statement has on the
transaction-local database state $\stl$ (i.e., how it transforms
$\stl$), and compose multiple such effects together to describe the
effect of the transaction as a whole. Importantly, this approach
generalizes to loops, where the effect of a loop can be inferred as a
well-defined function of the effect of its body, thanks to certain
pleasant properties enjoyed by the database programs modeled by
$\txnimp$.

@@

At the core of our approach is a simple language ($\SL$) to express
set transformations shown in Fig.~\ref{fig:logic-syntax}. The language
admits only set expressions that include variables ($x$), literals of
the form $\{x \,|\, \phi^0\}$ where $\phi^0$ is a propositional
(quantifier-free) formula on $x$, a restricted form of existential
quantification that binds a set $x$ satisfying proposition $\phi^0$ in
a set expression $s$, a monadic composition of two set expressions
($s_1$ and $s_2$) using bind ($\bind$), a conditional set expression
where the condition is a propositional formula, and a union of two set
expressions. Symbols $\stl$ and $\stg$ are also variables in $\SL$,
but are used to denote local and database states, respectively.  The
language is carefully chosen to be expressive enough to capture the
semantics of $\txnimp$ statements, yet simple enough to have a
semantics-preserving translation to the first-order logic.

@@

Fig.~\ref{fig:inference-rules} shows the syntax-directed inference
rules for $\txnimp$ commands inside a transaction $\C{TXN}_i$.
Inference depends on the $\I$-constrained rely relation $\R$, and the
high-level invariant $I$ for the reasons that will become clear
shortly.  Inference rules compute, for each command $c$, a (meta)
function $\F$ that returns a set as an expression in $\SL$, given a
pair of sets ($\stl$ and $\stg$) that describe local and a global
database states, respectively. The expression returned by
$\F(\stl,\stg)$ (abstractly) describes the set of records that get
added to $\stl$ as a result of executing $c$ under $\stl$ and $\stg$.
Thus, $\F$ captures the \emph{effect} part of the state transformer of
$c$, which is the function $\lambda(\stl,\stg).~\stl \cup
\F(\stl,\stg)$. For $\F$ to be useful in RG verification, it needs to
be stable w.r.t the rely relation $\R$. The stability condition on
effects is defined below:

@@

where, $\bar{v}$ are the variables that occur free in $\F$ (possible
because of how the inference rules are structured).  Intuitively, the
stability condition requires an effect to describe the same set of
records before and after the interference. The equivalence in $\SL$
translates to the equality in first-order logic, as will become clear
shortly. In the inference rules, stability is enforced constructively
by a meta function $\stabilize{\cdot}$, which accepts an effect and
returns a new effect that is guaranteed to be stable under $\R$.
$\stabilize{\cdot}$ achieves the stability guarantee by abstracting
away the bound $\stg$ in an unstable $\F$ to an existentially bound
$\stg'$ as described below:

@@

Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns an
$\F'$ that simply ignores its $\stg$ argument in favor of a generic

@@

hence only valid database states will ever be witnessed. From the

@@

\rulelabel{RG-Conseq} rule for transaction-bound commands (cf.  the
supplementary material). The weakening semantics chosen by

@@

\caption{A transaction that deposits an interest to a bank account}

@@

Here, an \C{add\_interest} transaction adds a positive interest to the
balance of a bank account, which is required to be positive ($I(\stg)
\Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge 0$). The transaction
starts by issuing a \C{select1} query, whose effect implicitly asserts
that there exists a record in the bank account database that is equal
to $a$ ($\exists(r\in\stg).~r=a$). However, this assertion is unstable

@@

update the account balance, and there will no longer exist an
$r\in\stg$ that is equal to $a$. Fortunately, the weakening operator
($\stabilize{\cdot}$) comes to the rescue here by weakening the
assertion to $\exists \stg'.I(\stg') \conj \exists(r\in\stg').~a=r$,
which is enough to prove that $\C{a.bal + pc*a.bal}\ge 0$, and verify
that \C{add\_interest} preserves the positive balance invariant.

The inference rules of SQL statements, like the RG rules, closely
follow the corresponding reduction rules in Fig.~\ref{fig:txnimp},
except that their language of expression is $\SL$. For instance, while
the reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the inference rule uses $\SL$'s bind operation to
\emph{compute} the set. Other SQL rules do likewise. The inference
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. For instance, the effect of a
sequence of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$
of $c_1$ and $c_2$, respectively, except that $F_2$ is applied to a
state ($\stl$) to which $F_1$ has already been applied, reflecting
their order of reduction. The inference rule for \C{FOREACH} takes
advantage of the $\SL$'s bind operator to lift the effect inferred for
the loop body to the level of the loop. Since records added to $\stl$
in each iteration of \C{FOREACH} are independent of the previous
iteration, sequential composition of the effects of different
iterations is same as their parallel composition. Since the loop body
is executed once per each $z\in x$, the effect of the the loop is the
union of effects ($\F$) for all $z\in x$, all applied to the same
state ($\stl$ and $\stg$). That is, $\F_{loop}(\stl,\stg) =
\bigcup_{z\in x}\F_{body}(\stl,\stg)$. From the definition of
set monad's bind, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which is what the rule infers.

@@

\begin{example}
Consider a database of couples that tracks the spouse
relationship. Each record has an $\idf$ field, a \C{name} field, and a
spouse id (\C{s\_id}) field. A $\txnimp$ transaction bearing id $i$
that populates the database with a given set ($xs$) of couples will
include the following code snippet:
\begin{smathpar}
\begin{array}{l}
  \foreache{xs}{(\lambda y.\lambda x.~ 
        \updatee{(\lambda z.~ \{\idf=z.\idf;\,\C{name}=z.\C{name};\,
                               \C{s\_id}=x.\idf\})}
                {(\lambda z.~ z.\idf = x.\C{s\_id})};\\
        \hspace*{1.27in} \inserte{x})}
\end{array}
\end{smathpar}
The effect inferred for the program, assuming no interference, is as
following:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~xs \bind 
    (\lambda x. \stg \bind 
      (\lambda z. \itel{z.\idf = x.\C{s\_id}}
                       {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
                               \C{s\_id}=x.\idf\}\}}
                       {\emptyset}) \\
      \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=false; \}\})
\end{array}
\end{smathpar}
Under the possibility of an interference affecting the stability, the
following stable effect is inferred ($I$ is the database invariant):
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~xs \bind 
    (\lambda x.~\existsl(\stg',I,\stg' \bind 
      (\lambda z. \itel{z.\idf = x.\C{s\_id}}
                       {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
                               \C{s\_id}=x.\idf\}\}}
                       {\emptyset})) \\
      \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=false; \}\})
\end{array}
\end{smathpar}
\end{example}

@@

The correspondence between the inference rules and the RG judgment of
Sec.~\ref{sec:reasoning} is stated thus\footnote{Proof can be found in
the supplementary material}:
\begin{theorem}
\label{thm:inference-sound}
  Forall i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
  then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
  I(\stg)}{~c~}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}
Where the set expression $\F(\emptyset,\stg)$ has a natural
interpretation, on the lines discussed in the next sub-section.
% Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
% \Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
% \R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by the
inference judgment. The soundness of a transaction's guarantee can now
be established w.r.t the effect $\F$ of the body. The
\rulelabel{RG-Txn} rule updated to use the inference judgment is shown
below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\\
  \R(\stl,\stg,\stg') \Leftrightarrow \I(\C{true},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  c \elabsto \F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \I(\C{false},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  \stable(\R_c,\F)\\
  \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Since the inference judgment $c\elabsto \F$ is fully automatic,
automating the application of the \rulelabel{RG-Txn} rule for a
transaction now requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $\I$ are formulas in
the first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

The algorithm ($\mssemof{\cdot}$) shown in Fig.~\ref{fig:logic}
encodes an $\SL$ expression in a FOL. Given a set expression $s$ of
$\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where $\phi$ encodes $s$ in
FOL, and $f$ is a boolean function that names the set defined by $s$.
Naming provides an easy handle to $s$, simplifying $\mssemof{\cdot}$.
Set variables ($x$, $\stl$, $\stg$, $\dots$) are encoded as
unconstrained boolean functions ($g_x$, $g_{\stl}$, $g_{\stg}$,
$\ldots$) that are named after the variable.  Encoding of a set
literal $\{a\,|\,\varphi\}$ introduces a new boolean function $g$ that
is $true$ for only those $a$ that satisfy $\varphi$.  Thus $g$
faithfully encodes the literal. The encoding of the $\existsl{}$
expression first eliminates the second-order quantification over sets.
Existentially bound set variable ($a$) is skolemized, i.e.,
instantiated with a new unbound variable $a'$, before encoding the
bound expression $s$ and the first-order constraint $\phi$. The
algorithm for encoding $\phi$, which is already a first-order formula,
is straightforward. It basically involves replacing set variables with
boolean functions named after the variables. The first-order encoding
for a bind expression describes the semantics of set monad's $\bind$
in FOL. Let $s_1$ be a set, and let $f$ be a function that maps each
variable in $s_1$ to a new set. Then, $s_2 = s_1 \bind f$ if and only
if for all $y\in s_2$, there exists an $x \in s_1$ such that $y =
f(x)$, and forall $x\in s_1$, $f(x)\in s_2$. The encoding adds new
constraints to this effect. Conditional set expressions and set union
are encoded straightforwardly.

Observe that the language of encoding shown in Fig.~\ref{fig:logic} is
a subset of logic that satisfies the following syntactic properties:
\begin{itemize}
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.  
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and boolean.
\end{itemize}
The fragment of FOL that satisfies the above syntactic conditions is
called Effectively Propositional first-order logic (denoted {\sf
EPR}), or Bernay-Sch{\"o}nfinkel-Ramsey (BSR) logic, which is known to
be decidable~\cite{z3epr}. The language of encoding, however, is a
combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
quantifier-free propositions ($\varphi$) that encode object language
expressions are drawn, and (b).  $\Prop^1$, the theory from which
invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
and state the following theorem:
\begin{theorem}
  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
{\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}, hence:
\begin{theorem}
  $\SL[{\sf SLA},{\sf EPR}]$ is decidable.
\end{theorem}
The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
Isolation specifications presented in Sec.~\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf EPR}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$
fragment, making the verification a practical exercise.

>>>>

The Rely-Guarantee framework presented in the previous section
facilitates modular proofs for weakly-isolated transactions, but
imposes a non-trivial annotation burden.  In particular, it requires
each statement ($c$) of the transaction to be annotated with a stable
pre- ($P$) and post-condition ($Q$), and loops to be annotated with
stable inductive invariants ($\psi$). While weakest pre-condition
style predicate transformers can help in inferring intermediate
assertions for regular statements, loop invariant inference remains
challenging, even for the simple form of loops considered here.  As an
alternative, we present an inference algorithm based on state
transformers.  The idea is to infer the logical effect that each
statement has on the transaction-local database state $\stl$ (i.e.,
how it transforms $\stl$), and compose multiple such effects together
to describe the effect of the transaction as a whole.  Importantly,
this approach generalizes to loops, where the effect of a loop can be
inferred as a well-defined function of the effect of its body, thanks
to certain pleasant properties enjoyed by the database programs
modeled by $\txnimp$.  Interpreting database semantics as functional
transformations on sets (described in terms of their logical effects)
enables an inference mechanism that can leverage off-the-shelf SMT
solvers for automated verification.

@@

At the core of our approach is a simple language ($\SL$) to express
set transformations (see \ref{fig:logic-syntax}). The language admits
only set expressions that include variables ($x$), literals of the
form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
(quantifier-free) formula on $x$, a restricted form of existential
quantification that binds a set $x$ satisfying proposition $\phi$ in
a set expression $s$, a monadic composition of two set expressions
($s_1$ and $s_2$) composed using a bind ($\bind$) operation, a
conditional set expression where the condition is a propositional
formula, and a union of two set expressions. Symbols $\stl$ and $\stg$
are also variables in $\SL$, but are used to denote local and database
states, respectively.  The language is carefully chosen to be
expressive enough to capture the semantics of $\txnimp$ statements
(and is representative of SQL semantics), yet simple enough to have a
semantics-preserving translation to a decidable fragment of
first-order logic.

@@

Fig.~\ref{fig:inference-rules} shows the syntax-directed state
transformer inference rules for $\txnimp$ commands inside a
transaction $\C{TXN}_i$.  Inference depends on the $\I$-constrained
rely relation $\R$, and the high-level invariant $I$ for the reasons
that will become clear shortly.  Inference rules compute, for each
command $c$, a (meta) function $\F$ that returns a set as an
expression in $\SL$, given a pair of sets ($\stl$ and $\stg$) that
describe local and a global database states, respectively. The
expression returned by $\F(\stl,\stg)$ (abstractly) describes the set
of records that get added to $\stl$ as a result of executing $c$ under
$\stl$ and $\stg$.  Thus, $\F$ captures the \emph{effect} part of the
state transformer of $c$, which is the function
$\lambda(\stl,\stg).~\stl \cup \F(\stl,\stg)$\footnote{Recall that the
  operational semantics treats deletion of records as the addition of
  the deleted record with its \C{del} field set to true in the local
  store.}. For $\F$ to be useful in RG verification, it needs to be
stable w.r.t the rely relation $\R$. The stability condition on
effects can be defined thus:

@@

where, $\bar{v}$ are the variables that occur free in $\F$; this is
possible because of how the inference rules are structured.
Intuitively, the stability condition requires an effect to describe
the same set of records before and after the interference. The
equivalence in $\SL$ translates to the equality in first-order logic,
as we describe below. In the inference rules, stability is
enforced constructively by a meta function $\stabilize{\cdot}$, which
accepts an effect and returns a new effect that is guaranteed to be
stable under $\R$.  $\stabilize{\cdot}$ achieves the stability
guarantee by abstracting away the bound $\stg$ in an unstable $\F$ to
an existentially bound $\stg'$ as described below:

@@

Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns a
transformer $\F'$ that simply ignores its $\stg$ argument in favor of a generic

@@

and hence only valid database states will ever be witnessed. From the

@@

\rulelabel{RG-Conseq} rule for transaction-bound commands.  The weakening semantics chosen by

@@

\caption{A transaction that deposits an interest to a bank account.}

@@

Here, an \C{add\_interest} transaction adds a positive interest
(\C{pc}) to the balance of a bank account, which is required to be
positive ($I(\stg) \Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge
0$). The transaction starts by issuing a \C{select1} query, whose
effect implicitly asserts that there exists a record in the bank
account database that is equal to $a$
($\exists(r\in\stg).~r=a$). However, this assertion is unstable

@@

update the account balance, so that such a record no longer exists in
the database.  Fortunately, the weakening operator
($\stabilize{\cdot}$) allows us to weaken the assertion to $\exists
\stg'.I(\stg') \conj \exists(r\in\stg').~a=r$ (Fig.~\ref{fig:logic}
formalizes the encoding of {\sf exists} to logic), which is enough to
prove that $\C{a.bal + pc*a.bal}\ge 0$, and verify that
\C{add\_interest} preserves the positive balance invariant.

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. For instance, the effect of a
sequence of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$
of $c_1$ and $c_2$, respectively, except that $F_2$ is applied to a
state ($\stl$) to which $F_1$ has already been applied, reflecting
their order of reduction. The inference rule for \C{FOREACH} takes
advantage of the $\SL$'s bind operator to lift the effect inferred for
the loop body to the level of the loop. Since records added to $\stl$
in each iteration of \C{FOREACH} are independent of the previous
iteration, sequential composition of the effects of different
iterations is same as their parallel composition. Since the loop body
is executed once per each $z\in x$, the effect of the the loop is the
union of effects ($\F$) for all $z\in x$, all applied to the same
state ($\stl$ and $\stg$). That is, $\F_{loop}(\stl,\stg) =
\bigcup_{z\in x}\F_{body}(\stl,\stg)$. From the definition of the set
monad's bind operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.

@@

%% \begin{example}
%% Consider a database of couples that tracks the spouse
%% relationship. Each record has an $\idf$ field, a \C{name} field, and a
%% spouse id (\C{s\_id}) field. A $\txnimp$ transaction bearing id $i$
%% that populates the database with a given set ($xs$) of couples will
%% include the following code snippet:
%% \begin{smathpar}
%% \begin{array}{l}
%%   \foreache{xs}{(\lambda y.\lambda x.~ 
%%         \updatee{(\lambda z.~ \{\idf=z.\idf;\,\C{name}=z.\C{name};\,
%%                                \C{s\_id}=x.\idf\})}
%%                 {(\lambda z.~ z.\idf = x.\C{s\_id})};\\
%%         \hspace*{1.27in} \inserte{x})}
%% \end{array}
%% \end{smathpar}
%% The effect captured by a state transformer inferred for the program, assuming no interference would be:
%% \begin{smathpar}
%% \begin{array}{l}
%%   \lambda(\stl,\stg).~xs \bind 
%%     (\lambda x. \stg \bind 
%%       (\lambda z. \itel{z.\idf = x.\C{s\_id}}
%%                        {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
%%                                \C{s\_id}=x.\idf\}\}}
%%                        {\emptyset}) \\
%%       \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=\mathit{false}; \}\})
%% \end{array}
%% \end{smathpar}
%% Under the possibility of an interference affecting the stability, the
%% following stable effect is inferred ($I$ is the database invariant):
%% \begin{smathpar}
%% \begin{array}{l}
%%   \lambda(\stl,\stg).~xs \bind 
%%     (\lambda x.~\existsl(\stg',I,\stg' \bind 
%%       (\lambda z. \itel{z.\idf = x.\C{s\_id}}
%%                        {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
%%                                \C{s\_id}=x.\idf\}\}}
%%                        {\emptyset})) \\
%%       \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=false; \}\})
%% \end{array}
%% \end{smathpar}
%% \end{example}

@@

The correspondence between the inference rules and the RG judgment of
Sec.~\ref{sec:reasoning} is stated thus\footnote{The proof can be found in
the supplementary material.}:
\begin{theorem}
\label{thm:inference-sound}
  \emph{Forall} i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
  then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
  I(\stg)}{~c~}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}
\noindent where the set expression $\F(\emptyset,\stg)$ has a natural
interpretation as discussed below.
% Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
% \Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
% \R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's guarantee can now
be established w.r.t the effect $\F$ of the body. The
\rulelabel{RG-Txn} rule so updated is shown
below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\\
  \R(\stl,\stg,\stg') \Leftrightarrow \I(\C{true},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  c \elabsto \F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \I(\C{false},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  \stable(\R_c,\F)\\
  \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $\I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

The algorithm ($\mssemof{\cdot}$) shown in Fig.~\ref{fig:logic}
encodes an $\SL$ expression in a decidable fragment of FOL. Given a
set expression $s$ of $\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where
$\phi$ encodes $s$ in FOL, and $f$ is a Boolean function that names
the set defined by $s$.  Naming provides an easy handle to $s$,
simplifying $\mssemof{\cdot}$.  Set variables ($x$, $\stl$, $\stg$,
$\dots$) are encoded as unconstrained Boolean functions ($g_x$,
$g_{\stl}$, $g_{\stg}$, $\ldots$) that are named after the variable.
Encoding of a set literal $\{a\,|\,\varphi\}$ introduces a new Boolean
function $g$ that is \emph{true} for only those $a$ that satisfy $\varphi$.
Thus $g$ faithfully encodes the literal. The encoding of the
$\existsl{}$ expression first eliminates the second-order
quantification over sets.  Existentially bound set variable ($a$) is
skolemized, i.e., instantiated with a new unbound variable $a'$,
before encoding the bound expression $s$ and the first-order
constraint $\phi$. The algorithm for encoding $\phi$, which is already
a first-order formula, is straightforward. It basically involves
replacing set variables with Boolean functions named after the
variables. The first-order encoding for a bind expression describes
the semantics of set monad's bind operator in FOL. Let $s_1$ be a set, and
let $f$ be a function that maps each variable in $s_1$ to a new
set. Then, $s_2 = s_1 \bind f$ if and only if for all $y\in s_2$,
there exists an $x \in s_1$ such that $y = f(x)$, and forall $x\in
s_1$, $f(x)\in s_2$. The encoding adds new constraints to this
effect. Conditional set expressions and set union are encoded
straightforwardly.

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a subset of logic that satisfies the following syntactic properties:
\begin{itemize}
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.  
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
\end{itemize}
This fragment of FOL known as EPR (Effectively Propositional logic)
is known to be decidable~\cite{z3epr}. The language of encoding, however, is a
combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
quantifier-free propositions ($\varphi$) that encode object language
expressions are drawn, and (b).  $\Prop^1$, the theory from which
invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
and state the following theorem:
\begin{theorem}
  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}, hence:
\begin{theorem}
  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
\end{theorem}
The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
The isolation specifications presented in Sec.~\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf EPR}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
making automated verification a practical exercise.

### 2017-07-07 related-work.tex
edits to inference

\vspace*{-4pt}
\paragraph{Inference.}  Prior work by Vafeiadis~\cite{Vaf10,Vaf10a}
describe \emph{action inference}, an inference procedure for computing
rely and guarantee relations in the conetxt of RGSep~\cite{VP07}, an
integration of rely-guarantee and separation logic~\cite{Rey02} that
allows one to precisely reason about local and shared state of a
concurrent program. The ideas underlying action inference have been
used to prove memory safety, linearizability, shape invariant
inference, etc.  of fine-grained concurrent data structures.  While
our motivation is similar to theirs (inference of intermediate
assertions and invariants), the context of study (transactions
vs. shared-memory concurrency), the objects being analyzed (relational
database tables vs. concurrent data structures), and the analysis
technique used to drive inference (state transformers vs. abstract
interpretation) are quite different.

>>>>

\vspace*{-4pt}
\paragraph{Inference.}  Prior work by Vafeiadis~\cite{Vaf10,Vaf10a}
describe \emph{action inference}, an inference procedure for computing
rely and guarantee relations in the context of RGSep~\cite{VP07}, an
integration of rely-guarantee and separation logic~\cite{Rey02} that
allows one to precisely reason about local and shared state of a
concurrent program. The ideas underlying action inference have been
used to prove memory safety, linearizability, shape invariant
inference, etc.  of fine-grained concurrent data structures.  While
our motivation is similar (automated inference of intermediate
assertions and local invariants), the context of study (transactions
vs. shared-memory concurrency), the objects being analyzed (relational
database tables vs. concurrent data structures), the properties being
verified (integrity constraints over relational tables vs. memory
safety, or linearizability of concurrent data strucgture operations)
and the analysis technique used to drive inference (state transformers
vs. abstract interpretation) are quite different.
